Open archaeology, open source?

### 2023-10-13 analysis/_03-data_methodology.qmd
Updated the descriptions and interpretations of findings.
Also some minor updates to branch/fork terminology.

```{r tbl-glossary}
#| tbl-cap: Glossary of git and GitHub terminology
tribble(
  ~term, ~definition,
  "Comment", "On GitHub, Text post attached to an issue, including the first one that describes the issue",
  "Commit", "Set of changes (addition, alteration, or deletion) to files in a repository that has been recorded by git as one entry in its log",
  "Commit access", "Ability to make changes to a repository directly, without making a pull request",
  "Contributor", "User that has made at least one commit to a specified repository",
  "Follow", "Add activity by another user to a user's timeline",
  "Forge", "Web-based platform for hosting, distributing and facilitating collaboration on version controlled computer code, e.g. GitHub, GitLab, Codeberg",
  "Fork", "Copy of a repository owned by another user; forking is a prerequisite to making a pull request",
  "git", "Open source version control software",
  "GitHub", "Commercial platform that freely hosts git repositories and provides extended collaboration and social networking features, such as pull requests, issues and stars",
  "GitLab", "Open source alternative to GitHub",
  "CodeBerg", "Open source alternative to GitHub",
  "Issue", "Feature of GitHub that records and tracks a bug report, feature request or other suggestion in a repository",
  "Maintainer", "Individual that has overall control of a repository, generally assumed to be its primary contributor. Repositories can have multiple users with commit access in addition to the maintainer.",
  "Merge", "Accept a pull request and incorporate its changes into a repository.",
  "Organization", "Entity representing a group of users, which can also own repositories",
  "Pull request", "Mechanism by which users that don't have commit access to a repository can contribute to it. The repository's maintainer or another user with commit access must decide whether to merge (accept) the changes, or decline them.",
  "Repository", "Individual project that uses git for version control. Can include a mix of different types of files.",
  "Star", "GitHub's version of a 'like', applied by users to a repository.",
  "Timeline", "Chronological feed of GitHub activity from repositories a user has starred and other users they follow. Also includes repositories that a user is not following if they are 'trending' or determined relevant by GitHub's algorithm.",
  "User", "On GitHub, an individual with an account that can own repositories",
  "Version control", "System for tracking changes (additions, alterations, or deletions) in a set of files, typically but not exclusively computer code"
) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase) |>
  fmt_markdown(definition)
```

We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, which are designed to facilitate open source software development by hosting public git repositories.
Git is a protocol designed to facilitate open and distributed contributions to a common code base. <!-- mention other version control systems like SVN and mercurial maybe? -->
It operates by providing mechanisms for synchronizing communal, web-based public repositories with local iterations stored on contributors' private workstations.
Contributors who volunteer or are assigned to develop, inspect, or revise a specific aspect of a code base download a copy of the public repository into their own work environment, create a branch in which they apply their modifications, and then request that their branch be merged into the central code base.
After a public repository's maintainers decide to merge the proposed changes into the communal code base, other developers may use git to download these changes while keeping their own independent branches intact.

@@

In addition to serving as hosting providers for public git repositories, platforms like as GitHub, GitLab and Codeberg support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.
They also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, identify certain repositories as "favourites" and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
However, these additional features are meant to complement and enhance the experience of contributing to open source projects, but are not actually part of the git protocol.
See @tbl-glossary for a glossary of terminology used to describe many of these concepts and features, and which we use throughout this paper.

>>>>

```{r tbl-glossary}
#| tbl-cap: Glossary of git and GitHub terminology
tribble(
  ~term, ~definition,
  "Comment", "On GitHub, text post attached to an issue, including the first one that describes the issue",
  "Commit", "Set of changes (addition, alteration, or deletion) to files in a repository that has been recorded by git as one entry in its log",
  "Commit access", "Ability to make changes to a repository directly, without making a pull request",
  "Contributor", "User that has made at least one commit to a specified repository",
  "Follow", "Add activity by another user to a user's timeline",
  "Forge", "Web-based platform for hosting, distributing and facilitating collaboration on version controlled computer code, e.g. GitHub, GitLab, Codeberg",
  "Fork", "Copy of a repository owned by another user; forking is a prerequisite to making a pull request",
  "git", "Open source version control software",
  "GitHub", "Commercial platform that freely hosts git repositories and provides extended collaboration and social networking features, such as pull requests, issues and stars",
  "GitLab", "Open source alternative to GitHub",
  "CodeBerg", "Open source alternative to GitHub",
  "Issue", "Feature of GitHub that records and tracks a bug report, feature request or other suggestion in a repository",
  "Maintainer", "Individual that has overall control of a repository, generally assumed to be its primary contributor. Repositories can have multiple users with commit access in addition to the maintainer.",
  "Merge", "Accept a pull request and incorporate its changes into a repository.",
  "Organization", "Entity representing a group of users, which can also own repositories",
  "Pull request", "Mechanism by which users that don't have commit access to a repository can contribute to it. The repository's maintainer or another user with commit access must decide whether to merge (accept) the changes, or decline them.",
  "Repository", "Individual project that uses git for version control. Can include a mix of different types of files.",
  "Star", "GitHub's version of a 'like', applied by users to a repository.",
  "Timeline", "Chronological feed of GitHub activity from repositories a user has starred and other users they follow. Also includes repositories that a user is not following if they are 'trending' or determined relevant by GitHub's algorithm.",
  "User", "On GitHub, an individual with an account that can own repositories",
  "Version control", "System for tracking changes (additions, alterations, or deletions) in a set of files, typically but not exclusively computer code"
) |>
  gt() |>
  cols_label_with(fn = tools::toTitleCase) |>
  fmt_markdown(definition)
```

We compiled the dataset by browsing collaborative software development platforms such as GitHub, GitLab and Codeberg, which are designed to facilitate open source software development by hosting public git repositories.
Git is a protocol designed to facilitate open and distributed contributions to a common code base. <!-- mention other version control systems like SVN and mercurial maybe? -->
It operates by providing mechanisms for synchronizing communal, web-based public repositories with local iterations stored on contributors' private workstations.
Contributors who volunteer or are assigned to develop, inspect, or revise a specific aspect of a code base download a copy of the public repository into their own work environment, create a fork in which they apply their modifications, and then request that their fork be merged into the central code base.
After a public repository's maintainers decide to merge the proposed changes into the communal code base, other developers may use git to download these changes while keeping their own independent forks intact.

@@

In addition to serving as hosting providers for public git repositories, platforms like GitHub, GitLab and Codeberg support common software developer and project management practices, such as issue and bug tracking, code-commenting, task management, identity and permissions management, web publishing, vulnerability detection, creation and maintenance of metadata, and financial sponsorship.
They also implement standard social media functions, like the ability to follow projects and individual users to receive updates on their activities, "star" certain repositories as favourites, and maintain a public-facing profile that includes personally-identifying information (e.g. profile picture, username, real name, employer or affiliation), references to all public activity on the platform, and links to the user's other social media profiles.
Code-sharing platforms thus serve as comprehensive developer portfolios and community networking resources.
However, these additional features are meant to complement and enhance the experience of contributing to open source projects, but are not actually part of the git protocol.
See @tbl-glossary for a glossary of terminology used to describe many of these concepts and features, and which we use throughout this paper.

### 2023-10-13 analysis/_04-open_archaeology.qmd
Updated the descriptions and interpretations of findings.
Also some minor updates to branch/fork terminology.

<!-- TODO: expand. -->
We found that many repositories do not explicitly denote an open source license, or any license at all [@tbl-repo-licenses].

>>>>

To enact their mandate of ensuring that anyone can access and modify software and other creative works, the open source and open science movements encourage developers and scientists to adopt open licenses.
Licenses are legally-bindings statements that stipulate how a creative work can be accessed and used.
Proprietary licenses usually require explicit permission to be granted so that the work can be accessed or modified, usually in exchange for financial compensation.
Open licenses, on the other hand, are more permissive, and allow anyone to use creative works without such harsh restrictions.
While it is certainly possible to write your own license, it is very common to simply use one of several standardized open licenses (see [choosealicense.com](https://choosealicense.com for a more comprehensive overview of various options)).
Some licenses, like GNU, MIT and Apache, are explicitly suited for distributing software, and specify certain use cases that are afforded by digital media.
Other licenses, like the Creative Commons variants, are more suited to other kinds of creative works like books, articles, movies, music, photographs, and websites.
The Creative Commons licenses also include clauses that cater to academic or creative sensibilities, such as requirements to attribute credit to the original authors, to restrict commercial use, and to propagate similar restrictions in derivative works.

We found that the projects that host their code on GitHub use a wide variety of standard licenses.
However, nearly half of all repositories <!-- TODO: encode the percentage --> do not explicitly denote an open source license, or any license at all [@tbl-repo-licenses].
Maintainers may have assumed that making the code available, without explicitly stating permitting use, is enough to allow unrestricted access to the repository's contents.
This is evocative of a similar ignorance among academics regarding the unsustainable use of GitHub as a long-term code and data hosting platform (@milliken2021; @escamilla2022; @escamilla2023).
<!-- TODO: Expand with additional info after seeing what is in the "other" category -->

### 2023-10-13 analysis/_05-collaboration.qmd
Updated the descriptions and interpretations of findings.
Also some minor updates to branch/fork terminology.

As well as hosting source code, GitHub and other software forges include systems for facilitating collaboration on code and other projects.
The basic collaborative workflow is inherited from git, which allows multiple users to commit (see @tbl-glossary for definitions of this and other git terminology used in this section) code to the repository.
A user with commit access to a repository can change any of its contents at will, so this is usually reserved for the project maintainer and known, trusted collaborators.
GitHub extends this model with its pull request feature, by which any user can fork a repository to which they don't have commit access, make changes, then offer to contribute those changes back to the original repository.
The maintainer can choose to merge (accept) or decline the pull request, facilitating contributions from a wider network of collaborators without the need for permission to be sought in advance.

@@

Another way GitHub users can engage with repositories and other users is with social media-like features such as starring a repository, commenting on an existing issue, or following a user.
These actions populate a timeline of through which users can see recent activity and discover new projects related to those they have interacted with in the past.^[This feature of GitHub's timeline was one of the primary ways we compiled open-archaeo.]
While not as a direct a contribution as pull requests or issues, these features can facilitate the formation and maintenance of collaborative networks, in the same way that other social media platforms serve other professional networks.
These features are more used more widely than forks, issues and pull requests (@fig-collab-features): `r percent(p_with_stars)` of repositories have at least one star and, in those repositories that use issues, `r percent(p_issues_with_comments)` of them has received at least one additional comment.

@@

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_50)`.
<!-- TODO: what do we make of this observation? -->

The prototypical collaborative open source project on GitHub has a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
<!-- TODO: note that we don't know if maintainers actually aimed for that model, or leave that for the discussion? -->
Only a small number of open-archaeo projects (<!-- TODO: do these share any characteristics in terms of platform or subject? -->) approach this model.
The majority of projects are short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.

>>>>

As well as hosting source code, GitHub and other software forges include systems for facilitating collaboration on code and other projects.
The basic collaborative workflow is inherited from git, which allows multiple users to commit code to the repository (see @tbl-glossary for definitions of this and other git terminology used in this section).
A user with commit access to a repository can change any of its contents at will, so this is usually reserved for the project maintainer and known, trusted collaborators.
GitHub extends this model with its pull request feature, by which any user can fork a repository to which they don't have commit access, make changes, then offer to contribute those changes back to the original repository.
The maintainer can choose to merge (accept) or decline the pull request, facilitating contributions from a wider network of collaborators without the need for permission to be sought in advance.

@@

Another way GitHub users can engage with repositories and other users is with social media-like features such as starring a repository, commenting on an existing issue, or following a user.
These actions populate a timeline of through which users can see recent activity and discover new projects related to those they have interacted with in the past.^[This feature of GitHub's timeline was one of the primary ways we compiled open-archaeo.]
While not as a direct a contribution as pull requests or issues, these features can facilitate the formation and maintenance of collaborative networks, in the same way that other social media platforms serve other professional networks.
These features are used more widely than forks, issues and pull requests (@fig-collab-features): `r percent(p_with_stars)` of repositories have at least one star and, in those repositories that use issues, `r percent(p_issues_with_comments)` of them has received at least one additional comment.

@@

Perhaps unsurprisingly, given the low uptake of GitHub's collaborative features, `r percent(p_one_contrib)` of open-archaeo repositories only contain commits from a single user.
Even in the minority of projects that have more than one contributor, work (as measured by number of recorded commits) is distributed highly unevenly (@fig-contributions).
The lead maintainer almost always does the lion share of the work: they are responsible for more than half of commits in `r percent(p_big_boy_50)` of projects and more than three-quarters in `r percent(p_big_boy_50)`.
This may be attributed to the steep learning curve commonly attributed to working with git.
While git can be a great way to track changes and manage distributed contributions to a common code base, it can also be unwieldy in situations when multiple users (especially those with less experience using git for collaborative purposes) are expected to contribute within short spans of time.
This meshes with our prior observations that projects which tend to exhibiting higher commit rates have less contributors.
Additionally, our analyses neglects to account for contributions that are not tracked via git or GitHub.
Those who do not code may provide creative guidance or feedback during in-person meetings, via email, or using alternative online messaging or social media platforms.
A more focused qualitative assessment of these non-coding and supportive work practices would shed more light on the totality of effort that goes into producing and maintaining open source projects.

<!-- TODO: Abrupt transition, need to revise -->
The prototypical and idyllic collaborative open source project ingrained into the common consciousness hypothetically comprises a core group of developers (often a single maintainer) that regularly commit new code, a wider network of collaborators that contribute through forks and pull requests, plus an active user base that create and comment on issues, who have indicated their support for the project by starring its repository.
It is unclear whether archaeological software developers actually aim to operate following this model, or whether it is even suitable for supporting what open science aims to achieve.
However it is clear that only a small number of open-archaeo projects (<!-- TODO: do these share any characteristics in terms of platform or subject? -->) operate according to this model.
The majority of projects are in fact short-lived, with few contributors and a small number of commits.
Use of GitHub's collaboration features is also generally low (@fig-collab-features), although the data also shows a divergence between the uptake of features that facilitate direct code contributions (forks, issues, pull requests), which have markedly zero-skewed distributions, versus more indirect, social media-like features (comments, stars), which are moderately well-used.
We hypothesise that this shows a preference for passive/reactive rather than active/proactive engagement with others work – a point we will return to in the conclusion.
